A Complete Algebra for File Synchronization

### 2015-11-07 2015/paper.tex
Detailed two sets of axioms. Style improvements.

\usepackage{alltt}
\usepackage{comment}
\usepackage{multicol}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{setspace}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{boxedminipage}
\usepackage{afterpage}

\newcommand{\setv}{\mathcal{V}}
\newcommand{\setvx}[1]{\mathcal{V}_{#1}}
\newcommand{\setf}{\setvx{F}}
\newcommand{\setd}{\setvx{D}}
\newcommand{\setb}{\setvx{\empt}}
\newcommand{\setp}{\mathcal{P}}
\newcommand{\empt}{\bot}
\newcommand{\parent}{\mathtt{parent}}
\newcommand{\FS}{\Phi} % {\mathrm{FS}}
\newcommand{\GS}{\Xi} % another filesystem
\newcommand{\pp}{\hat{p}} % parent path
\newcommand{\np}{\mathrel{\ooalign{$p$\cr\hidewidth\hbox{$-$}\cr}}} % path not related to p

\newcommand{\cbrk}{\mathtt{break}}
\newcommand{\fscommand}[2]{{#1#2}}
\newcommand{\cbb}{\fscommand{\empt\empt}}
\newcommand{\cbf}{\fscommand{\empt{F}}}
\newcommand{\cbd}{\fscommand{\empt{D}}}
\newcommand{\cfb}{\fscommand{F\empt}}
\newcommand{\cff}{\fscommand{FF}}
\newcommand{\cfd}{\fscommand{FD}}
\newcommand{\cdb}{\fscommand{D\empt}}
\newcommand{\cdf}{\fscommand{DF}}
\newcommand{\cdd}{\fscommand{DD}}
\newcommand{\cxy}{\fscommand{XY}}
\newcommand{\czw}{\fscommand{ZW}}

\title{Real-Life Algebraic File Synchronization}

@@

Moreover, in $\setv$ we distinguish between files ($\setf$) and directories ($\setd$), that is,
if $\setb = \{\empt\}$ then:
\[ \setv = \setb \cup \setf \cup \setd \]
The tree propety can then be defined as
\[ \forall \FS : \forall p\in\setp : \FS(p) \neq \empt \wedge \parent(p) \neq \empt \Rightarrow \FS(\parent(p)) \in \setd \]

% diagram
% /> D ---> F ---> BOT <\
% \--/               \--/

@@

We therefore propose to have a command for each pair of types.
For want of a better system, we will simply name our commands by concatenating
two of $\fscommand{D}{}$, $\fscommand{F}{}$ and $\fscommand{\empt}{}$, 
where the left sign notes the type of value
in the filesystem before the command is applied, and the right sign notes the type
afterwards. For example $mkdir(p,D)$ is $\cbd(p,D)$ and $rmdir(p)$ is $\cdb(p)$.

@@

So that we could reason about sequences of commands that break every filesystem
in the algebra, we introduce the command $\cbrk$ that simply breaks every filesystem.
We note that a command or a sequence of commands $S$ is applied to a filesystem
by prefixing the command or sequence to it:
\begin{itemize}
\item $S\FS$
\item $\cbrk\FS=\empt$
\item $\cff(p,F)\FS = \GS$, for which $\GS(p)=F$ holds
and $\GS(q)=\FS(q)$ for any path $q\neq p$.
\end{itemize}

@@

Pairs of commands in general will have the form
\[ \cxy(p,V_Y); \czw(q,V_W) \]
where $X,Y,Z,W\in\{D,F,\empt\}$, $p,q\in\setp$, and values are of the appropriate type: 
$V_Y\in\setvx{Y}$ and $V_W\in\setvx{W}$
(although commands where $Y=\empt$ or $W=\empt$ do not require a value).

@@

\begin{itemize}
\item $p=q$. If the two commands affect the same path, the pair of commands
will clearly behave differently than if the paths are different.
\item Separate paths, that is, $\parent(p)\neq q$ and $\parent(q)\neq p$. If the paths
are not directly related, then modifying the filesystem at one path is not going
to affect the outcome of the command on the other path.
The two commands will therefore commute and the pair will be reversible.
\item Directly related paths, that is, $\parent(p)=q$ or $\parent(q)=p$.
Some of these command pairs will not be reversible.
For example, $\cbd(\parent(p),D);\cbf(p,F)$ will work on some filesystems,
while $\cbf(p,F);\cbd(\parent(p),D)$ will break all filesystems.
\end{itemize}

One may think that we may be able to specify more about the reversibility
or possible simplification of a pair of commands in the last group if we also
knew whether the child path $p$ was the only (non-empty) child of its parent.
However, it is easy to see that encoding this information in a pair
is not going to enrich our algebra as, for example,
the behaviour of a hypothetical $\cfb(p_{\mathrm{only\ child}}); \cdb(\parent(p))$
is exactly the same as that of the more generic
$\cfb(p); \cdb(\parent(p))$. The former will break any filesystem in which
$p$ is not the only child as it violates its preconditions, but so will
the generic pair.

In summary, if we adopt the notation $\pp := \parent(p)$ and use $\np$ for a path not directly related to $p$,
then we have pairs of commands
of the following forms:
\begin{align*}
\cxy(p,V_Y)&; \czw(p,V_W) \\
\cxy(p,V_Y)&; \czw(\np,V_W) \\
\cxy(p,V_Y)&; \czw(\pp,V_W) \\
\cxy(\pp,V_Y)&; \czw(p,V_W)
\end{align*}

@@

>>>>

\usepackage{alltt}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{setspace}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{afterpage}
\usepackage{centernot}

\usepackage{tikz}

\newcommand{\setv}{\mathcal{V}}
\newcommand{\setvx}[1]{\mathcal{V}_{#1}}
\newcommand{\setf}{\setvx{F}}
\newcommand{\setd}{\setvx{D}}
\newcommand{\setb}{\setvx{\empt}}
\newcommand{\setp}{\mathcal{P}}
\newcommand{\empt}{\bot}
\newcommand{\parent}{\mathtt{parent}}
\newcommand{\FS}{\Phi} % {\mathrm{FS}}
\newcommand{\GS}{\Xi} % another filesystem
\newcommand{\pp}{p^\uparrow} % parent path
\newcommand{\np}{p_{\centernot\leftrightarrow}} % path not related to p

\newcommand{\cbrk}{\mathtt{break}}
\newcommand{\fscommand}[2]{{#1#2}}
\newcommand{\fsregcommandchar}[1]{\mathtt{#1}}
\newcommand{\fsregcommand}[2]{\fscommand{\fsregcommandchar{#1}}{\fsregcommandchar{#2}}}
\newcommand{\cbb}{\fsregcommand{\empt}{\empt}}
\newcommand{\cbf}{\fsregcommand{\empt}{F}}
\newcommand{\cbd}{\fsregcommand{\empt}{D}}
\newcommand{\cfb}{\fsregcommand{F}{\empt}}
\newcommand{\cff}{\fsregcommand{F}{F}}
\newcommand{\cfd}{\fsregcommand{F}{D}}
\newcommand{\cdb}{\fsregcommand{D}{\empt}}
\newcommand{\cdf}{\fsregcommand{D}{F}}
\newcommand{\cdd}{\fsregcommand{D}{D}}
\newcommand{\cxy}{\fscommand{X}{Y}}
\newcommand{\cyz}{\fscommand{Y}{Z}}
\newcommand{\cxz}{\fscommand{X}{Z}}
\newcommand{\czw}{\fscommand{Z}{W}}

\newcommand{\eqext}{\sqsubseteq}
\newcommand{\eqnrw}{\sqsupseteq}

\newcommand{\emptyseq}{[\,]}

\title{A Complete Algebra for File Synchronization}

@@

Moreover, in $\setv$ we distinguish between files ($\setf$) and directories ($\setd$), that is,
if $\setb = \{\empt\}$ then:
\[ \setv = \setb \cup \setf \cup \setd \]
The tree propety can then be defined as
\[ \forall \FS, p\in\setp : \FS(p) \neq \empt \wedge \parent(p) \neq \empt \Rightarrow \FS(\parent(p)) \in \setd \]

This means that as we move from a path to its child, its grandchild, and so on,
the types of values we encounter in the filesystem can only change according to the following
transition diagram:

% diagram
% /> D ---> F ---> BOT <\
% \--/               \--/
\input{type-transition}

In other words, the types will form a sequence like $D,D,\cdots,D,F,\empt,\empt,\cdots$.

@@

We therefore propose to have a command for each pair of types.
For want of a better system, we will simply name our commands by concatenating
two of $\fsregcommandchar{D}$, $\fsregcommandchar{F}$ and $\fsregcommandchar{\empt}$, 
where the left sign notes the type of value
in the filesystem before the command is applied, and the right sign notes the type
afterwards. For example $mkdir(p,D)$ is $\cbd(p,D)$ and $rmdir(p)$ is $\cdb(p)$.

@@

So that we could reason about sequences of commands that break every filesystem
in the algebra, we introduce the command $\cbrk$ that simply breaks every filesystem.
We note that a sequence of commands $S$ is applied to a filesystem
by prefixing the sequence to it:
\begin{itemize}
\item $S\FS$
\item $[\cbrk]\FS=\empt$
\item $[\cff(p,F)]\FS = \GS$, for which $\GS(p)=F$ holds
and $\GS(q)=\FS(q)$ for any path $q\neq p$.
\end{itemize}

@@

Pairs of commands in general will have the form
\[ [\cxy(p,V_Y); \czw(q,V_W)] \]
where $X,Y,Z,W\in\{\fsregcommandchar{D},\fsregcommandchar{F},\empt\}$, $p,q\in\setp$, and values are of the appropriate type: 
$V_Y\in\setvx{Y}$ and $V_W\in\setvx{W}$
(although commands where $Y=\empt$ or $W=\empt$ do not require a value).

@@

\begin{itemize}
\item $p=q$. If the two commands affect the same path, the pair of commands
will clearly behave differently than if the paths are different.
%
\item Separate paths, that is, $\parent(p)\neq q$ and $\parent(q)\neq p$. If the paths
are not directly related, then modifying the filesystem at one path is not going
to affect the outcome of the command on the other path.
The two commands will therefore commute and the pair will be reversible.
We will denote a path separate from $p$ with $\np$.
%
\item Directly related paths, that is, $\parent(p)=q$ or $\parent(q)=p$.
For simplicity, we will write $\pp$ for $\parent(p)$.
Some of these command pairs will not be reversible.
For example, $[\cbd(\pp,D);\cbf(p,F)]$ will work on some filesystems,
while $[\cbf(p,F);\cbd(\pp,D)]$ will break all filesystems.
\end{itemize}

One may think that we may be able to specify more about the reversibility
or possible simplification of a pair of commands in the last group if we also
knew whether the child path $p$ was the only (non-empty) child of its parent.
However, it is easy to see that encoding this information in a pair
is not going to enrich our algebra as, for example,
the behaviour of a hypothetical $[\cfb(p_{\mathrm{only\ child}}); \cdb(\pp)]$
is exactly the same as that of the more generic
$[\cfb(p); \cdb(\pp)]$. The former will break any filesystem in which
$p$ is not the only child as it violates its preconditions, but so will
the generic pair.

In summary, we have pairs of commands
of the following forms:
\begin{itemize}
\item[i.] \( [\cxy(p,V_Y); \czw(\np,V_W)] \)
\item[ii.] \( [\cxy(p,V_Y); \czw(p,V_W)] \)
\item[iii.] \( [\cxy(\pp,V_Y); \czw(p,V_W)] \)
\item[iv.] \( [\cxy(p,V_Y); \czw(\pp,V_W)] \)
\end{itemize}

%% What does it mean to restrict the inspected filesystems to a given relationship?

\section{Defining the Algebra}

In order to reason about sequences of commands, we define \emph{algebraic laws} on
commands. We use formal logic to build a proof system which is sound and
complete for its intended interpretation.

Similarly to \cite{NREC:alg}, we have two kinds of judgements
for sequences \(S_1\) and \(S_2\):
\begin{itemize}
\item \(S_1\equiv S_2\), or \emph{\(S_1\) is algebraically equivalent to
\(S_2\)}. Its intended interpretation is that they act in the same way on all
filesystems.
\item \(S_1\eqext S_2\), or \emph{\(S_2\) extends \(S_1\)}; its intended
interpretation is that they act in the same way on all
filesystems that \(S_1\) does not break.
\end{itemize}

We have two inference rules for the two judgements,
which make it possible to derive new statements from the axioms or other statements.
These are:

\emph{For any sequences \(S_1, S_2, S, S'\):}
\begin{itemize}
\item if \(S_1\equiv S_2\) then \(S;S_1;S'\equiv S;S_2;S'\)
\item if \(S_1\eqext S_2\) then \(S;S_1;S'\eqext S;S_2;S'\)
\item if \(S_1\equiv S_2\) then \(S_1\eqext S_2\).
\end{itemize}

Finally, we have a set of axioms or laws that describe the behaviour
of each command pair. We exclude commands that either break the
filesystem or leave it in the same state ($\cbb$) and for
simplicity's sake we merged axioms where the outcome is predictable
from the relationship between the paths in the commands.

\begin{itemize}
\item[i.] If the two paths in the pair are separate (not directly related), then
the commands commute with an equivalence:
\[ [\cxy(p,V_Y); \czw(\np,V_W)] \equiv [\czw(\np,V_W); \cxy(p,V_Y)] \]
%
\item[ii.] If the two paths are the same ($[\cxy(p,V_Y); \czw(p,V_W)]$), then
the pair will break every filesystem if $Z\ne Y$, that is, if the second
command expects a type of value different from the one left by the first command.
Otherwise, the pair will simplify to a single or no commands
if both $X$ and $W$ are $\fsregcommandchar{\empt}$ or $\fsregcommandchar{D}$:
\begin{gather*}
% bf=fb Sp
           [\cbf(p, F); \cfb(p)] \eqext \emptyseq \\
% bd=db Sp
           [\cbd(p, D); \cdb(p)] \eqext \emptyseq \\
% db=bd Sp
           [\cdb(p); \cbd(p, D)] \eqext [\cdd(p, D)] \\
% df=fd Sp
           [\cdf(p, F); \cfd(p, D)] \eqext [\cdd(p, D)]
\end{gather*}
In all other cases the commands are equivalent to a single command
replacing the input of the first command with the output of the second:
\[ [\cxy(p, V_Y); \cyz(p, V_Z)] \equiv [\cxz(p, V_Z)] \]
%
\item[iii.]
Of the pairs in which the first command is applied to the parent of the path
of the second command there are pairs to which no law applies.
We call these ``construction'' pairs:
\begin{gather*}
% bd>bf
           [\cbd(\pp, D); \cbf(p, F)] \\
% bd>bd
           [\cbd(\pp, D_1); \cbd(p, D_2)] \\
% fd>bf
           [\cfd(\pp, D); \cbf(p, F)] \\
% fd>bd
           [\cfd(\pp, D_1); \cbd(p, D_2)]
\end{gather*}
Otherwise commands in these pairs commute iff the first command is $\cdd$;
all other pairs break every filesystem. For example:
\begin{gather*}
[\cdd(\pp, D); \cbf(p, F)] \equiv [\cbf(p, F); \cdd(\pp, D)] \\
[\cdf(\pp, F); \cbd(p, D)] \equiv \cbrk
\end{gather*}
%
\item[iv.]
The final group of pairs are the ones in which the path in the second
command is the parent of the one in the first.
These behave similarly to the previous group. There are
so-called ``destruction'' pairs to which no law applies:
\begin{gather*}
% fb<db ??
           [\cfb(p); \cdb(\pp)] \\
% fb<df ??
           [\cfb(p); \cdf(\pp, F)] \\
% db<db ??
           [\cdb(p); \cdb(\pp)] \\
% db<df ??
           [\cdb(p); \cdf(\pp, F)]
\end{gather*}
Otherwise the commands commute iff the \emph{second} command is $\cdd$.
All other pairs break every filesystem.
\end{itemize}

\section{Real-Life Variant}

In real-life applications, we often do not want to consider metadata stored in
directories (\cite{BZ}). In other words, the contents of directories are all equal,
which can be modelled by assuming that $|\setd|=1$.
Note that this creates a convenient symmetry where there are only a single value
for each type, $D$ and $\empt$, that can be repeated as we move from a path to its child.

This also means that $\cdd$, like $\cbb$, becomes a command that either breaks a filesystem
or leaves it in the same state, and we will disregard it in our algebra.
The axioms will simplify as follows:

\begin{itemize}
\item[i.] Pairs with separate paths are still reversible,
but \emph{these are the only pairs that are so}.
%
\item[ii.] If the two paths are the same ($[\cxy(p,V_Y); \czw(p,V_W)]$), then, as above,
the pair will break every filesystem if $Z\ne Y$.
Otherwise, the pair is extended by an empty sequence
if both $X$ and $W$ are $\fsregcommandchar{\empt}$ or $\fsregcommandchar{D}$:
\begin{gather*}
% bf=fb Sp
           [\cbf(p, F); \cfb(p)] \eqext \emptyseq \\
% bd=db Sp
           [\cbd(p, D); \cdb(p)] \eqext \emptyseq \\
% db=bd Sp
           [\cdb(p); \cbd(p, D)] \eqext \emptyseq \\
% df=fd Sp
           [\cdf(p, F); \cfd(p, D)] \eqext \emptyseq
\end{gather*}
And in all other cases, as above, the commands are equivalent to a single command
replacing the input of the first command with the output of the second:
\[ [\cxy(p, V_Y); \cyz(p, V_Z)] \equiv [\cxz(p, V_Z)] \]
%
\item[iii.]
Of the pairs in which the first command is applied to the parent path,
the same ``construction'' pairs still have no laws, but all other pairs
break all filesystems.
%
\item[iv.]
And we still have the ``destruction'' pairs among the pairs where the second command
is applied to the parent path, and all other pairs break all filesystems.
\end{itemize}

@@

\bibitem{BZ} Bill Zissimopoulos, ...
